### 2015-03-19 swmpr_manu.tex
intro and methods beef up, figure captions, started tables, added lots of refs

\subsection*{Data retrieval}

SWMP data can be used in R after they are obtained directly from the \ac{CDMO} through an online query or by using the retrieval functions provided in this package.  In the latter case, the IP address for the computer making the request must be registered with \ac{CDMO}.  This can be done by following instructions \href{http://cdmo.baruch.sc.edu/webservices.cfm}{here}.  The \href{http://cdmo.baruch.sc.edu/data/metadata.cfm}{metadata} should also be consulted for available data, including the parameters and date ranges for each monitoring station.  Metadata are included as a .csv file with data requested from the \ac{CDMO} and can also be obtained using the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (individual site) functions.  Again, these functions will only work if the computer's IP address is registered with \ac{CDMO}.

@@

\subsection*{An overview of methods for swmpr objects}

Three categories of functions are available: retrieve, organize, and analyze.  The retrieval functions import the data into R as a swmpr object for use with the organize and analyze functions.  Methods defined for swmpr objects can be applied with the organize and analyze functions.  These methods are available for generic functions specific to this package, in addition to methods for existing generic functions available from other packages.  S3 methods are implemented in all cases.

The organize functions are used to clean or prepare the data for analysis, including removal of QAQC flags, subsetting, creating a standardized time series vector, and combining data of different types.  The \texttt{qaqc} function is a simple screen to retain values from the data with specified QAQC flags, described \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{here}.  Each parameter in the swmpr data typically has a corresponding QAQC column of the same name with the added prefix \texttt{f\_}.  Values in the QAQC column specify a flag from -5 to 5.  Generally, only data with the  \texttt{0} QAQC flag should be used, which is the default option for the \texttt{qaqc} function.  Data that do not satisfy QAQC criteria are converted to NA values.   Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. Processed data will have QAQC columns removed, in addition to removal of values in the actual parameter columns that do not meet the criteria.

@@

Viewing the number of observations for each parameter that are assigned to a QAQC flag may be useful for deciding how to process the data with ), \texttt{qaqc}.  The \texttt{qaqcchk} function can be used to view this information.  Consult the \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{online documentation} for a description of each QAQC flag.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# view the number observations in each QAQC flag}
\hlkwd{qaqcchk}\hlstd{(dat)}
\end{alltt}
\end{kframe}
\end{knitrout}

Raw nutrient data obtained from the \ac{CDMO} will usually include replicate samples that were taken within a few minutes of each other.  The \texttt{rem\_reps.swmpr} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.  The datetimestamp column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defauls to the mean.  Other suggested functions include the \texttt{median}, \texttt{min}, or \texttt{max}.  The entire function call including treatment of NA values should be passed to the \texttt{FUN} argument (see the examples).  The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if QAQC columns are present.

@@

Time series can be smoothed to better characterize a signal independent of noise.  Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.  The \texttt{smoother} function can be used to smooth parameters in a swmpr object using a specified window size.  This method is a simple wrapper to \texttt{filter}.  The \texttt{window} argument specifies the number of observations included in the moving average.  The \texttt{sides} argument specifies how the average is calculated for each observation (see the documentation for \texttt{filter}).  A value of 1 will filter observations within the window that are previous to the current observation, whereas a value of 2 will filter all observations within the window centered at zero lag from the current observation. As before, the \texttt{params} argument specifies which parameters to smooth.

@@

\hlcom{# plot to see the difference}
\hlkwd{plot}\hlstd{(do_mgl} \hlopt{~} \hlstd{datetimestamp,} \hlkwc{data} \hlstd{= dat,} \hlkwc{type} \hlstd{=} \hlstr{'l'}\hlstd{)}
\hlkwd{lines}\hlstd{(test,} \hlkwc{select} \hlstd{=} \hlstr{'do_mgl'}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{'red'}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{)}
\end{alltt}
\end{kframe}

{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-15}

@@

@@

A common issue with any statistical analysis is the treatment of missing values.  Missing data can be excluded from the analysis, included but treated as true zeroes, or interpolated based on similar values.  In either case, an analyst should have a strong rationale for the chosen method.  A common approach used to handle missing data in time series analysis is linear interpolation.  A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.  A challenge with linear interpolation is an appropriate gap size for fitting missing observations.  The ability of the interpolated data to approximate actual trends is a function of the gap size.  Interpolation between larger gaps are less likely to resemble patterns of an actual parameter, whereas interpolation between smaller gaps are more likely to resemble actual patterns.  An appropriate gap size limit depends on the unique characteristics of specific datasets or parameters.  The \texttt{na.approx} function can be used to interpolate gaps in a swmpr object.  A required argument for the function is \texttt{maxgap} which defines the maximum gap size  for interpolation.

@@

\hlcom{# plot for comparison}
\hlkwd{par}\hlstd{(}\hlkwc{mfrow} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{3}\hlstd{,} \hlnum{1}\hlstd{))}
\hlkwd{plot}\hlstd{(do_mgl} \hlopt{~} \hlstd{datetimestamp, dat,} \hlkwc{main} \hlstd{=} \hlstr{'Raw'}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{'l'}\hlstd{)}
\hlkwd{plot}\hlstd{(do_mgl} \hlopt{~} \hlstd{datetimestamp, test,} \hlkwc{col} \hlstd{=} \hlstr{'red'}\hlstd{,}
  \hlkwc{main} \hlstd{=} \hlstr{'Interpolation - maximum gap of 10 records'}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{'l'}\hlstd{)}
\hlkwd{lines}\hlstd{(dat,} \hlkwc{select} \hlstd{=} \hlstr{'do_mgl'}\hlstd{)}
\hlkwd{plot}\hlstd{(do_mgl} \hlopt{~} \hlstd{datetimestamp, test2,} \hlkwc{col} \hlstd{=} \hlstr{'red'}\hlstd{,}
  \hlkwc{main} \hlstd{=} \hlstr{'Interpolation - maximum gap of 30 records'}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{'l'}\hlstd{)}
\hlkwd{lines}\hlstd{(dat,} \hlkwc{select} \hlstd{=} \hlstr{'do_mgl'}\hlstd{)}
\end{alltt}
\end{kframe}

{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-16}

@@

@@

The \texttt{decompose} function requires a ts object with a specified frequency as input.  The \texttt{decomp} function converts the input swmpr vector to a ts object prior to \texttt{decompose}.  This requires an explicit input defining the frequency of the parameter in the time series.  For example, the frequency of a parameter with diurnal periodicity would be 96 if the time step is 15 minutes (4 * 24).  The frequency of a parameter with seasonal periodicity would be 35040 (4 * 24 * 365).  For simplicity, character strings of \texttt{'daily'} or \texttt{'seasonal'} can be supplied in place of numeric values.  A starting value of the time series must be supplied in the latter case.  Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.

Note that the \texttt{decompose} function is a relatively simple approach and alternative methods should be investigated if a more sophisticated decomposition is desired.

@@

\hlcom{# decomposition and plot}
\hlstd{test} \hlkwb{<-} \hlkwd{decomp}\hlstd{(dat,} \hlkwc{param} \hlstd{=} \hlstr{'do_mgl'}\hlstd{,} \hlkwc{frequency} \hlstd{=} \hlstr{'daily'}\hlstd{)}
\hlkwd{plot}\hlstd{(test)}
\end{alltt}
\end{kframe}

{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-17}

@@

@@

An alternative approach to time series decomposition is provided by the \texttt{decomp\_cj} function, which is a simple wrapper to the \texttt{decompTs} function in the wq package.  Theory describing this method is described in \cite{Cloern10}.  The function is similar to \texttt{decomp.swmpr} with a few key differences.  The \texttt{decomp.swmpr} function decomposes the time series into a trend, seasonal, and random component, whereas the current function decomposes into the grandmean, annual, seasonal, and events components.  For both functions, the random or events components, respectively, can be considered anomalies that don't follow the trends in the remaining categories.  The \texttt{decomp\_cj} function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.  This approach is meant for nutrient data that are obtained on a monthly cycle.  The function will also work with continuous water quality or weather data but note that the data are first aggregated on the monthly scale before decomposition.  Accordingly, short-term variation less than one-month will be removed. Additional arguments passed to \texttt{decompTs} can be used with \texttt{decomp\_cj}, such as \texttt{startyr}, \texttt{endyr}, and \texttt{type}.  Values passed to \texttt{type} are \texttt{mult} (default) or \texttt{add}, referring to multiplicative or additive decomposition.  See the documentation for \texttt{decompTs} for additional explanation and examples.

@@

\hlcom{# decomposition of chl, ggplot}
\hlkwd{decomp_cj}\hlstd{(dat,} \hlkwc{param} \hlstd{=} \hlstr{'chla_n'}\hlstd{)}
\end{alltt}
\end{kframe}

{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-18}

@@

@@

A reserve map with all stations can be obtained using the \texttt{map\_reserve} function.  This function is a simple wrapper to functions in the ggmap package. The current function is limited to Google maps, which allows four map types that can be set with the \texttt{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.  The \texttt{zoom} argument may have to be chosen through trial and error depending on the spatial extent of the reserve.  See the help documentation for ggmap for more info on zoom.  Additionally, station locations are returned using the \texttt{site\_codes\_ind} function if the computer making the request has the IP address registered with \ac{CDMO}. Otherwise, a local and possibly outdated file is used.  Use the contact at the \ac{CDMO} \href{http://cdmo.baruch.sc.edu/webservices.cfm}{web services} to register your IP.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# plot the stations at Jacques Cousteau reserve}
\hlkwd{map_reserve}\hlstd{(}\hlstr{'jac'}\hlstd{)}
\end{alltt}
\end{kframe}

{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-19}

@@

Several graphics showing seasonal and annual trends for a given SWMP parameter can be obtained using the \texttt{plot\_summary} function.  The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.  Anomalies are defined as the difference between the monthly or annual average from the grand mean for the parameter.  Monthly anomalies are in relation to the grand mean for the same month across all years.  All data are aggregated for quicker plotting.  Nutrient data are based on monthly averages, whereas weather and water quality data are based on daily averages.  Cumulative precipitation data are based on the daily maximum. The function returns a graphics object (Grob) of multiple ggplot objects.  An interactive Shiny widget that uses this function is available (see the \nameref{supp_info}).

\hlcom{## plot}
\hlkwd{plot_summary}\hlstd{(dat,} \hlkwc{param} \hlstd{=} \hlstr{'chla_n'}\hlstd{)}
\end{alltt}
\end{kframe}

{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-20}

@@

The following is an example that shows how to use the function from a combined water quality and weather data set.  The results can be plotted using \texttt{plot\_metab}.

@@

\hlcom{## estimate metabolism}
\hlstd{res} \hlkwb{<-} \hlkwd{ecometab}\hlstd{(dat,} \hlkwc{trace} \hlstd{=} \hlnum{FALSE}\hlstd{)}
\hlkwd{plot_metab}\hlstd{(res)}
\end{alltt}
\end{kframe}

{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-21}

@@

@@

See help documentation for more details on each function (e.g., \texttt{?all\_params}).

\paragraph{Retrieve}

\texttt{all\_params} Retrieve up to 100 records starting with the most recent at a given station, all parameters.  Wrapper to \texttt{exportAllParamsXMLNew} function on web services.

\texttt{all\_params\_dtrng} Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter.  Maximum of 1000 records. Wrapper to \texttt{exportAllParamsDateRangeXMLNew}.

\texttt{import\_local} Import files from a local path.  The files must be in a specific format, specifically those returned from the \ac{CDMO} using the [zip downloads](http://cdmo.baruch.sc.edu/aqs/zips.cfm) option for a reserve.

\texttt{single\_param} Retrieve up to 100 records for a single parameter starting with the most recent at a given station.  Wrapper to \texttt{exportSingleParamXMLNew} function on web services.

@@

\texttt{comb.swmpr} Combines swmpr objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station. Only different data types can be combined.

\texttt{qaqc.swmpr} Remove QAQC columns and remove data based on QAQC flag values for a swmpr object.  Only applies if QAQC columns are present.

\texttt{qaqcchk.swmpr} View a summary of the number of observations in a swmpr object that are assigned to different QAQC flags used by \ac{CDMO}.  The output is used to inform further processing but is not used explicitly.

>>>>

\subsection*{SWMP overview and data retrieval}

Four core data elements are collected through the \ac{SWMP} monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.  The SWMPr package is developed for the continous abiotic monitoring network which includes a majority of the data collected by \ac{SWMP}.  Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).  Each reserve has no less than four water quality stations and one weather station at fixed locations.  Water quality and weather data are collected at 15 minute intervals, whereas nutrient data are collected monthly at each water quality station.  All data are made accessible through the \ac{CDMO} web portal, where multiple \ac{QAQC} measures are used to screen the information.  The final data include all observations with relevant \ac{QAQC} flags indicating the appropriate qualifier (\href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{view codes}).

The \ac{CDMO} web portal was established to support priority areas of \ac{SWMP} that focus on the continuation and advancement of data and information management.  As such, \ac{CDMO} provides access to over 35 million water quality, weather, and nutrient records that have been authenticated through systematic \ac{QAQC} procedures.  Estuary data must be obtained from \ac{CDMO} prior to using most of the functions within the SWMPr package.  In most cases, the analysis needs will typically define the location, date range, and parameters of interest that need to be obtained from the \ac{CDMO}.  All stations in the \ac{SWMP} network are identified by a 7 or 8 character name that specifies the reserve, station, and parameter type.  For example, `apaebwq' is the water quality identifier for the East Bay station at the Apalachicola reserve.  Similarly, a suffix of `met' or `nut' would specify the weather (meteorological) or nutrients station.

\ac{SWMP} data can be used in R after they are obtained directly from the \ac{CDMO} through an online query or by using the retrieval functions provided in this package.  Prior to any data request, the site, parameter type, and date ranges need to be identified based on the analysis needs.  All reserve names, stations, and date ranges for the water quality, weather, or nutrients data can be viewed on the \ac{CDMO} \href{http://cdmo.baruch.sc.edu/}{website}. Alternatively, the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (single site) functions provided by the SWMPr package can be used to view the same information.  As noted below, the computer's IP address must be registered by \ac{CDMO} staff before using the function.

\subsection*{Structure of the SWMPr package}

The SWMPr package was developed by considering a standard workflow that categorizes the functions as one of three steps based on their intended use: {\bf{\it retrieving}}, {\bf{\it organizing}}, and {\bf{\it analyzing}}.  Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class.  Functions for organizing and analyzing the data provide methods for working with the \texttt{swmpr} object class.  An additional class of functions, termed `miscellaneous', are also included as helpers for the main functions.  The following describes a general approach for using each category of functions based on a standard data workflow.

\paragraph{Data retrieval}

\ac{SWMP} data must be obtained from the \ac{CDMO} prior to organizing and analyzing the information.  Two approaches can be used.  First, the data can be obtained outside of R using the online query forms on the \ac{CDMO} website (\href{http://cdmo.baruch.sc.edu/get/landing.cfm}{see here}).  Second, functions from the SWMPr package can be used to import the data directly from the online server (\Cref{tab:retrieve}).  In the latter case, the IP address for the computer making the request must be registered with \ac{CDMO}.  This can be done by following instructions \href{http://cdmo.baruch.sc.edu/webservices.cfm}{here}.  The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with \ac{CDMO}.

% table for retrieval funcitons
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Data retrieval functions available from the SWMPr package.",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:retrieve",     table.env = FALSE)%
\begin{table}[!tbp]
\caption{Data retrieval functions available from the SWMPr package.\label{tab:retrieve}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\texttt{all\_params}&Retrieve up to 100 records starting with the most recent at a given station, all parameters.  Wrapper to \texttt{exportAllParamsXMLNew} function on web services.\tabularnewline
\texttt{all\_params\_dtrng}&Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter.  Maximum of 1000 records. Wrapper to \texttt{exportAllParamsDateRangeXMLNew}.\tabularnewline
\texttt{import\_local}&Import files from a local path.  The files must be in a specific format, specifically those returned from the \ac{CDMO} using the zip downloads option for a reserve.\tabularnewline
\texttt{single\_param}&Retrieve up to 100 records for a single parameter starting with the most recent at a given station.  Wrapper to \texttt{exportSingleParamXMLNew} function on web services.\tabularnewline
\texttt{site\_codes}&Metadata for all stations, wrapper to \texttt{exportStationCodesXMLNew} function on web services.\tabularnewline
\texttt{site\_codes\_ind}&Metadata for all stations at a single site, wrapper  to \texttt{NERRFilterStationCodesXMLNew} function on web services.\tabularnewline
\hline
\end{tabular}\end{center}

\end{table}

@@

Example data as raw, comma-separated files have not been included in the package due to size limitations.  However, a sample dataset can be \href{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}{downloaded} for use with the examples below.  This dataset has an identical format as the data returned from the zip downloads feature of the \ac{CDMO}.  Processed verions of the raw data are included with the package as binary data files (.RData) to decrease processing times with the examples.  Information for each binary file can be viewed as follows.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# view help files for complementary data}
\hlcom{# all files are samples from Apalachicola Bay}

\hlcom{# cat point station, nutrients}
\hlopt{?}\hlstd{apacpnut}

\hlcom{# cat point station, water quality}
\hlopt{?}\hlstd{apacpwq}

\hlcom{# dry bar station, water quality}
\hlopt{?}\hlstd{apadbwq}

\hlcom{# east bay station, weater}
\hlopt{?}\hlstd{apaebmet}
\end{alltt}
\end{kframe}
\end{knitrout}

\paragraph{Data organizing}

The retrieval functions import the data into R as a swmpr object for use with the organize and analyze functions. The organize functions are used to clean or prepare the data for analysis, including removal of QAQC flags, subsetting, creating a standardized time series vector, and combining data of different types.

The \texttt{qaqc} function is a simple screen to retain values from the data with specified QAQC flags (described \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{here}).  Each parameter in the \texttt{swmpr} data typically has a corresponding QAQC column of the same name with the added prefix \texttt{f\_}.  Values in the QAQC column specify a flag from -5 to 5.  Generally, only data with the  \texttt{0} \ac{QAQC} flag should be used, which is the default option for the \texttt{QAQC} function.  Data that do not satisfy \ac{QAQC} criteria are converted to \texttt{NA} values.   Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. Processed data will have QAQC columns removed, in addition to removal of values in the actual parameter columns that do not meet the criteria.

@@

Viewing the number of observations for each parameter that are assigned to a \ac{QAQC} flag may be useful for deciding how to process the data with ), \texttt{qaqc}.  The \texttt{qaqcchk} function can be used to view this information.  Consult the \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{online documentation} for a description of each QAQC flag.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# view the number of observations in each QAQC flag}
\hlkwd{qaqcchk}\hlstd{(dat)}
\end{alltt}
\end{kframe}
\end{knitrout}

Raw nutrient data obtained from the \ac{CDMO} will usually include replicate samples that were taken within a few minutes of each other.  The \texttt{rem\_reps.swmpr} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.  The \texttt{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defauls to the mean.  Other suggested functions include the \texttt{median}, \texttt{min}, or \texttt{max}.  The entire function call including treatment of NA values should be passed to the \texttt{FUN} argument (see the examples).  The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if QAQC columns are present.

@@

Time series can be smoothed to better characterize a signal independent of noise (\Cref{fig:smooth_ex}).  Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.  The \texttt{smoother} function can be used to smooth parameters in a \texttt{swmpr} object using a specified window size.  This method is a simple wrapper to \texttt{filter}.  The \texttt{window} argument specifies the number of observations included in the moving average.  The \texttt{sides} argument specifies how the average is calculated for each observation (see the documentation for \texttt{filter}).  A value of 1 will filter observations within the window that are previous to the current observation, whereas a value of 2 will filter all observations within the window centered at zero lag from the current observation. As before, the \texttt{params} argument specifies which parameters to smooth.  See \Cref{fig:smooth_ex} for the output from the code.

@@

\hlcom{# plot to see the difference}
\hlkwd{plot}\hlstd{(do_mgl} \hlopt{~} \hlstd{datetimestamp,} \hlkwc{data} \hlstd{= dat,} \hlkwc{type} \hlstd{=} \hlstr{'l'}\hlstd{)}
\hlkwd{lines}\hlstd{(test,} \hlkwc{select} \hlstd{=} \hlstr{'do_mgl'}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{'red'}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{)}
\end{alltt}
\end{kframe}\begin{figure}[!ht]

{\centering \includegraphics[width=\maxwidth]{figure/smooth_ex}

@@

\caption[Raw and smoothed dissolved oxygen data for a two-week period after using the \texttt{smoother} function]{Raw and smoothed dissolved oxygen data for a two-week period after using the \texttt{smoother} function.\label{fig:smooth_ex}}
\end{figure}

@@

A common issue with any statistical analysis is the treatment of missing values.  Missing data can be excluded from the analysis, included but treated as true zeroes, or interpolated based on similar values.  In either case, an analyst should have a strong rationale for the chosen method.  A common approach used to handle missing data in time series analysis is linear interpolation.  A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.  A challenge with linear interpolation is an appropriate gap size for fitting missing observations.  The ability of the interpolated data to approximate actual trends is a function of the gap size.  Interpolation between larger gaps are less likely to resemble patterns of an actual parameter, whereas interpolation between smaller gaps are more likely to resemble actual patterns.  An appropriate gap size limit depends on the unique characteristics of specific datasets or parameters.  The \texttt{na.approx} function can be used to interpolate gaps in a \texttt{swmpr} object.  A required argument for the function is \texttt{maxgap} which defines the maximum gap size for interpolation.  See \Cref{fig:interp_ex} for the output from the following code.

@@

\hlcom{# plot for comparison}
\hlkwd{par}\hlstd{(}\hlkwc{mfrow} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{3}\hlstd{,} \hlnum{1}\hlstd{))}
\hlkwd{plot}\hlstd{(do_mgl} \hlopt{~} \hlstd{datetimestamp, dat,} \hlkwc{main} \hlstd{=} \hlstr{'Raw'}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{'l'}\hlstd{)}
\hlkwd{plot}\hlstd{(do_mgl} \hlopt{~} \hlstd{datetimestamp, test,} \hlkwc{col} \hlstd{=} \hlstr{'red'}\hlstd{,}
  \hlkwc{main} \hlstd{=} \hlstr{'Interpolation - maximum gap of 10 records'}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{'l'}\hlstd{)}
\hlkwd{lines}\hlstd{(dat,} \hlkwc{select} \hlstd{=} \hlstr{'do_mgl'}\hlstd{)}
\hlkwd{plot}\hlstd{(do_mgl} \hlopt{~} \hlstd{datetimestamp, test2,} \hlkwc{col} \hlstd{=} \hlstr{'red'}\hlstd{,}
  \hlkwc{main} \hlstd{=} \hlstr{'Interpolation - maximum gap of 30 records'}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{'l'}\hlstd{)}
\hlkwd{lines}\hlstd{(dat,} \hlkwc{select} \hlstd{=} \hlstr{'do_mgl'}\hlstd{)}
\end{alltt}
\end{kframe}\begin{figure}[!ht]

{\centering \includegraphics[width=\maxwidth]{figure/interp_ex}

@@

\caption[Examples illustrating use of the \texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.]{Examples illustrating use of the \texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.\label{fig:interp_ex}}
\end{figure}

@@

The \texttt{decompose} function requires a \texttt{ts} object with a specified frequency as input.  The \texttt{decomp} function converts the input \texttt{swmpr} vector to a \texttt{ts} object prior to \texttt{decompose}.  This requires an explicit input defining the frequency of the parameter in the time series.  For example, the frequency of a parameter with diurnal periodicity would be 96 if the time step is 15 minutes (4 * 24).  The frequency of a parameter with seasonal periodicity would be 35040 (4 * 24 * 365).  For simplicity, character strings of \texttt{`daily'} or \texttt{`seasonal'} can be supplied in place of numeric values.  A starting value of the time series must be supplied in the latter case.  Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.  Note that the \texttt{decompose} function is a relatively simple approach and alternative methods should be investigated if a more sophisticated decomposition is desired.  \Cref{decomp_ex1} is an example of the \texttt{decomp} function.

@@

\hlcom{# decomposition and plot}
\hlstd{test} \hlkwb{<-} \hlkwd{decomp}\hlstd{(dat,} \hlkwc{param} \hlstd{=} \hlstr{'do_mgl'}\hlstd{,} \hlkwc{frequency} \hlstd{=} \hlstr{'daily'}\hlstd{)}
\hlkwd{plot}\hlstd{(test)}
\end{alltt}
\end{kframe}\begin{figure}[!ht]

{\centering \includegraphics[width=\maxwidth]{figure/decomp_ex1}

@@

\caption[An additive decomposition of dissolved oxygen into a trend, seasonal, and random component using the \texttt{decomp} function]{An additive decomposition of dissolved oxygen into a trend, seasonal, and random component using the \texttt{decomp} function.\label{fig:decomp_ex1}}
\end{figure}

@@

An alternative approach to time series decomposition is provided by the \texttt{decomp\_cj} function, which is a simple wrapper to the \texttt{decompTs} function in the wq package.  Theory describing this method is provided by Cloern and Jassby \cite{Cloern10}.  The function is similar to \texttt{decomp.swmpr} with a few key differences.  The \texttt{decomp.swmpr} function decomposes the time series into a trend, seasonal, and random component, whereas the current function decomposes into the grandmean, annual, seasonal, and events components.  For both functions, the random or events components, respectively, can be considered anomalies that do not follow the trends in the remaining categories.  The \texttt{decomp\_cj} function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.  This approach works best for nutrient data that are typically obtained on a monthly cycle.  The function will also work with continuous water quality or weather data but note that the data must first aggregated on the monthly scale before decomposition.  Additional arguments passed to \texttt{decompTs} can be used with \texttt{decomp\_cj}, such as \texttt{startyr}, \texttt{endyr}, and \texttt{type}.  Values passed to \texttt{type} are \texttt{mult} (default) or \texttt{add}, referring to multiplicative or additive decomposition.  \Cref{decomp_ex2} shows the results from the \texttt{decomp\_cj} function applied to a multi-year chlorophyll time series.

@@

\hlcom{# decomposition of chl, ggplot}
\hlkwd{decomp_cj}\hlstd{(dat,} \hlkwc{param} \hlstd{=} \hlstr{'chla_n'}\hlstd{)}
\end{alltt}
\end{kframe}\begin{figure}[!ht]

{\centering \includegraphics[width=\maxwidth]{figure/decomp_ex2}

@@

\caption[Additive decomposition of a multi-year chlorophyll time series into the grandmean, annual, seasonal, and events components using the \texttt{decomp\_cj} function]{Additive decomposition of a multi-year chlorophyll time series into the grandmean, annual, seasonal, and events components using the \texttt{decomp\_cj} function.\label{fig:decomp_ex2}}
\end{figure}

@@

Several graphics showing seasonal and annual trends for a given SWMP parameter can be obtained using the \texttt{plot\_summary} function.  The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.  Anomalies are defined as the difference between the monthly or annual average from the grand mean for the parameter.  Monthly anomalies are in relation to the grand mean for the same month across all years.  All data are aggregated for quicker plotting.  Nutrient data are based on monthly averages, whereas weather and water quality data are based on daily averages.  Cumulative precipitation data are based on the daily maximum. The function returns a graphics object (Grob) of multiple ggplot objects.  An interactive Shiny application \cite{shiny15} that uses this function is available (see the \nameref{supp_info}).

\hlcom{## plot}
\hlkwd{plot_summary}\hlstd{(dat,} \hlkwc{param} \hlstd{=} \hlstr{'chla_n'}\hlstd{)}
\end{alltt}
\end{kframe}\begin{figure}[!ht]

{\centering \includegraphics[width=\maxwidth]{figure/summary_ex}

@@

\caption[Summaries of a multi-year chlorophyll time series using the \texttt{plot\_summary} function]{Summaries of a multi-year chlorophyll time series using the \texttt{plot\_summary} function.  Summaries include monthly distributions (means on top left, quantiles on bottom left), monthly histograms (center), monthly means by year (top right), deviation from monthly means (middle right), and annual trends as deviations from the grand mean (bottom right)\label{fig:summary_ex}}
\end{figure}

@@

The following is an example that shows how to use the function from a combined water quality and weather data set.  The results can be plotted using \texttt{plot\_metab} (\Cref{metab_ex}).

@@

\hlcom{## estimate metabolism}
\hlstd{res} \hlkwb{<-} \hlkwd{ecometab}\hlstd{(dat,} \hlkwc{trace} \hlstd{=} \hlnum{FALSE}\hlstd{)}
\hlkwd{plot_metab}\hlstd{(res)}
\end{alltt}
\end{kframe}\begin{figure}[!ht]

{\centering \includegraphics[width=\maxwidth]{figure/metab_ex}

@@

\caption[Monthly aggregations of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data at Apalachicola Bay, Florida]{Monthly aggregations of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data at Apalachicola Bay, Florida.\label{fig:metab_ex}}
\end{figure}

\paragraph{Miscellaneous functions}

Several additional functions are provided that cannot be described by the above categories.  These functions are generally used within the main functions but may be useful for more customized evaluation of \ac{SWMP} data.

Finally, a reserve map with all stations can be obtained using the \texttt{map\_reserve} function.  This function is a simple wrapper to functions in the ggmap package \cite{ggmap13}. The current function is limited to Google maps, which allows four map types that can be set with the \texttt{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.  The \texttt{zoom} argument may have to be chosen through trial and error depending on the spatial extent of the reserve.  See the help documentation for ggmap for more info on zoom.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# plot the stations at Jacques Cousteau reserve}
\hlkwd{map_reserve}\hlstd{(}\hlstr{'jac'}\hlstd{)}
\end{alltt}
\end{kframe}\begin{figure}[!ht]

{\centering \includegraphics[width=\maxwidth]{figure/map_ex}

\caption[Locations of all sites at the Jacques Cousteau reserve using the \texttt{map\_reserve} function]{Locations of all sites at the Jacques Cousteau reserve using the \texttt{map\_reserve} function.\label{fig:map_ex}}
\end{figure}

@@

@@

\texttt{comb.swmpr} Combines \texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station. Only different data types can be combined.

\texttt{qaqc.swmpr} Remove QAQC columns and remove data based on QAQC flag values for a \texttt{swmpr} object.  Only applies if QAQC columns are present.

\texttt{qaqcchk.swmpr} View a summary of the number of observations in a \texttt{swmpr} object that are assigned to different QAQC flags used by \ac{CDMO}.  The output is used to inform further processing but is not used explicitly.
